---
title: "Multi-Omics Integration Analysis - MOFA"
author: "Adam El Idrissi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Practical overview

This R Markdown document preserves the workflow and execution order of the supplied MOFA.R script, applied in the same dataset we analysed previously. This document contains detailed descriptions of the code cells and the resulting outputs in relation to the dataset used in this practical. It provides an overview of the multi-omics integration process performed using `Multi-Omics Factor Analysis (MOFA)`. The official MOFA2 tutorials, available at `https://biofam.github.io/MOFA2/tutorials.html`, which provides additional examples of model training and downstream analysis.

MOFA utilizes Matrix Factorization framework for the unsupervised integration of multiple omics datasets. It learns an interpretable, low-dimensional representation of the integrated data through a set of latent factors. These factors capture the major sources of variation across the analysed omics modalities and can subsequently be examined in relation to sample groups, biological characteristics, and influential molecular features.

------------------------------------------------------------------------

# 1. Package installation and loading

The required packages are obtained from both CRAN and Bioconductor repositories, with `MOFA2` installed through `BiocManager`. `MOFA2` is the main package used for multi-omics integration in this practical. `basilisk` is used to manage the Python environment and dependencies required by the MOFA2 implementation. In this workflow, this environment is accessed explicitly when the model is trained using `run_mofa(..., use_basilisk = TRUE)`. The stable Bioconductor installation of MOFA2 uses `basilisk` to configure the required Python environment automatically. `edgeR` is used to process the RNA-seq count matrix before integration. `data.table`, `dplyr`, and `tidyr` are loaded for dataframes manipulation.

```{r package-loading}


########## MOFA ################################################################

## Load all the required libraries 
# BiocManager::install("MOFA2")


library(MOFA2)
library(data.table)
library(ggplot2)
library(dplyr)
library(psych)
library(GGally)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(basilisk)
library(edgeR)

```

------------------------------------------------------------------------

# 2. Load the RNA-seq, methylation, and metadata files

Boar taint is one of the most extensively studied phenotypes in swine production. It is characterized by an unpleasant urine- or feces-like odor and flavour that is released during the cooking of meat from uncastrated male pigs. Based on the intensity of this phenotype, the animals included in this study were classified into three groups: Low, Medium, and High boar taint.

Our two datasets originate from the 3 groups of boars, differentiated by their taints (Low - Medium - High taint), where one dataset contains Liver RNA-seq measurements and the other contains Methylation measurements. The RNA matrix consists of 75,375 genes measured across 9 boars (3 from each phenotypes), whereas the methylation matrix contains 100,000 methylation observations measured from the same 9 boars (3 from each phenotypes). For vertical integration using the MOFA framework, it is essential that the same biological samples are shared across all omics modalities. In our case, the nine boars were profiled at different molecular levels, allowing the integration of transcriptomic and methylation information from the same individuals.

```{r data-import}
##### load all the required files and arrange the files for MOFA Analysis
RNA_Seq_Data=read.csv("RNA_Seq_Liver.csv", row.names = 1)
head(RNA_Seq_Data)[1:5]


Methylation=read.csv("Epigenetic_matrix_filtered_100000.csv", row.names = 1)
rownames(Methylation)=Methylation$region
Methylation$X=NULL
Methylation$region =NULL

head(Methylation)

metadat_columns=read.csv("Metadata.csv")
head(metadat_columns)
metadat_columns$X=NULL

```

------------------------------------------------------------------------

# 3. Metadata filtering and sample alignment

The metadata file contains the identifiers of each boar together with its corresponding boar taint phenotype (Low, Medium, or High). In this step, the metadata are prepared to be compatible with the MOFA integration framework. One important concept during multi-omics integration is `shared samples measurements`. This means that the samples measurements from the individuals must be compatible across all other modalities. For example, if Dataset A contains the samples identifiers as ID1, ID2, ID3, then Dataset B and Dataset C must also contain the same identifier (ID1, ID2, ID3). Any mismatch in the sample naming would result in non-functioning data fusion, and consequently biasing the results outcome.

```{r metadata-filtering}
###################### Metadata Filtering ######################################
# Add prefix based on boar taint group
metadat_columns$SampleName <- ifelse(
  grepl("high", metadat_columns$Group, ignore.case = TRUE), 
  paste0("H", metadat_columns$SampleID),
  ifelse(
    grepl("medium", metadat_columns$Group, ignore.case = TRUE), 
    paste0("M", metadat_columns$SampleID),
    ifelse(
      grepl("low", metadat_columns$Group, ignore.case = TRUE), 
      paste0("L", metadat_columns$SampleID),
      paste0("X", metadat_columns$SampleID)  # fallback if none match
    )
  )
)
metadat_columns$Group <- sub(" .*", "", metadat_columns$Group)



# Create the old and new column name mapping
metadat_columns$OldSampleName <- paste0("X", metadat_columns$SampleID)  # Original names
name_map <- setNames(metadat_columns$SampleName, metadat_columns$OldSampleName)  # Named vector: old -> new
# Rename methylation matrix columns
common_meth_cols <- intersect(colnames(Methylation), names(name_map))
colnames(Methylation)[match(common_meth_cols, colnames(Methylation))] <- name_map[common_meth_cols]
# Rename RNA-seq matrix columns
common_rna_cols <- intersect(colnames(RNA_Seq_Data), names(name_map))
colnames(RNA_Seq_Data)[match(common_rna_cols, colnames(RNA_Seq_Data))] <- name_map[common_rna_cols]
# Update metadata rownames to match new names
rownames(metadat_columns) <- metadat_columns$SampleName
# Filter metadata to keep only samples present in your omics data
omics_samples <- colnames(Methylation)  
sample_metadata <- metadat_columns[metadat_columns$SampleName %in% omics_samples, ]


# Double-check that sample orders match
all(colnames(RNA_Seq_Data) == colnames(Methylation))
# Reorder columns of methylation matrix to match RNA-seq
meth_matrix <- Methylation[, colnames(RNA_Seq_Data)]
####### Check that they're now aligned
all(colnames(meth_matrix) == colnames(RNA_Seq_Data))  # should return TRUE

## Change the colnames of metadata
colnames(sample_metadata)[colnames(sample_metadata) == "SampleName"] <- "sample"
colnames(sample_metadata)[colnames(sample_metadata) == "Group"] <- "group"

```

------------------------------------------------------------------------

# 4. Normalize and filter the RNA-seq data

Before integrating the transcriptomic data with any other data modality, the RNA-seq counts must be normalized to create a stable distribution. The raw count matrix are initially pre-processed by removing lowly expressed genes using `filterByExpr()`. Normalization factors are then estimated with `calcNormFactors()` for batch to batch estimation and false positive/negative results, and the counts are transformed into log<sub>2</sub> Counts Per Million (CPM) values using a prior count to avoid undefined values produced by log(0) (mathematically undefined real number). The final processed RNA-seq data leaves with 9 boars and the top 5000 most variable genes.

```{r rna-normalization}
########### Normalize and filter for downstream analysis #######################

# Create DGEList object
dge <- DGEList(counts = RNA_Seq_Data)
keep <- filterByExpr(dge)  
dge <- dge[keep, , keep.lib.sizes=FALSE]
# Normalize with TMM
dge <- calcNormFactors(dge)
# Log2-CPM transformation
log_RNA <- cpm(dge, log = TRUE, prior.count = 1)
# Filter top 3000 most variable genes
gene_var <- apply(log_RNA, 1, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:5000]
log_RNA_filtered <- log_RNA[top_genes, ]

dim(gene_var)
dim(log_RNA_filtered)

```

------------------------------------------------------------------------

# 5. Transform and filter the methylation data

Before integrating the methylation data with any other data modalities, in the same process of RNA-seq, methylation data is subject to pre-processing. The data are first converted into a numeric matrix, after which the methylation β-values are transformed into M-values using a logit transformatiion. This transformation is presented as log2 ratio transformation of the intensities methylated probes vs unmethylated probes. The reference of this transformation is in the following paper `https://pmc.ncbi.nlm.nih.gov/articles/PMC3012676/`. To improve data quality, methylation sites with more than 20% missing values are removed, and samples exceeding the same missingness threshold are also excluded. The final outcome of the methylation dataset is 5000 methylated regions in 9 candidate boars.

```{r methylation-filtering}
### Methylation data
# Ensure beta values are in [0, 1]
Methylation[Methylation <= 0] <- 0
Methylation[Methylation >= 1] <- 1 
summary(Methylation)
# Convert data frame to numeric matrix
Methylation <- as.matrix(Methylation)
summary(Methylation)
storage.mode(Methylation) <- "numeric"  # make sure all values are numeric
# Convert to M-values
Mval <- log2(Methylation / (1 - Methylation))
# Replace Inf and -Inf with NA safely
Mval[is.infinite(Mval)] <- NA
class(Mval)
# keep features with <20% NA
clean_meth <- Mval[apply(Mval, 1, function(x) sum(is.na(x)) < length(x) * 0.2), ]  
clean_meth <- clean_meth[, colSums(is.na(clean_meth)) < nrow(clean_meth) * 0.2]

# Calculate variance per CpG (row), ignoring NAs
var_per_cpg <- apply(clean_meth, 1, var, na.rm = TRUE)
# Sort CpGs by variance decreasing and get top 5000 CpGs
top_CPGs <- names(sort(var_per_cpg, decreasing = TRUE))[1:5000]
# Subset methylation data to top variable CpGs
Meth_view <- clean_meth[top_CPGs, ]
# Check dimensions
dim(Meth_view)
# Confirm shared sample names
common_samples <- intersect(colnames(log_RNA_filtered), colnames(Meth_view))



```

------------------------------------------------------------------------

# 6. Prepare the multi-omics data views

Before constructing the MOFA model, the RNA-seq and methylation datasets must contain the same biological samples in the same column order. The object common_samples, defined in the previous step, contains the sample identifiers shared by both omics datasets. The RNA-seq matrix is therefore subset using log_RNA_filtered[, common_samples], while the methylation matrix is subset using Meth_view[, common_samples]. This operation has two purposes. First, it removes any sample that is not present in both datasets. Second, it ensures that corresponding columns in the two matrices refer to the same biological individual. For example, the first column of RNA_view and the first column of Meth_view must represent the same sample. This alignment is essential because MOFA integrates information measured across the same set of samples.

The two processed datasets are then stored in the named list multi_omics_data. Each element of this list represents a separate omics view: - RNAseq contains the filtered and normalized RNA-seq data. - Methylation contains the filtered methylation M-values.

The function as.matrix() converts each object into a numeric matrix before it is supplied to MOFA. In both matrices, rows represent molecular features and columns represent samples. The RNA-seq matrix therefore contains genes in rows, whereas the methylation matrix contains CpG regions or methylation features in rows. It is important to note that the datasets are not merged into a single dataframe. Instead, they remain as separate matrices within the same list. This preserves the identity of each omics modality and allows MOFA to model variation within and across the RNA-seq and methylation views.

```{r multi-omics-data}
# Subset and reorder both datasets to same sample order
RNA_view <- log_RNA_filtered[, common_samples]
Meth_view <- Meth_view[, common_samples]



# Build multi-omics data list (9 sample names in columns, rows have RNASeq in one object, 
# similarly in a separate object another list with the CpG data)- They are NOT merged in a single object or data frame.
multi_omics_data <- list(
  RNAseq = as.matrix(RNA_view),
  Methylation = as.matrix(Meth_view)
)

head(multi_omics_data$RNAseq)
head(multi_omics_data$Methylation)


```

------------------------------------------------------------------------

# 7. Create, prepare, and train the MOFA model

## Construction and training of the MOFA model

The two aligned omics views are used to construct the initial MOFA object with `create_mofa()`. The resulting object contains the RNA-seq and methylation matrices together with the sample-group assignments supplied through `groups = sample_metadata$group`. At this stage, the model has not yet been trained; the function only organizes the input data into the structure required by MOFA2. The `plot_data_overview()` function provides a visual summary of the input structure. It provides the number of datasets that we're currently willing to train, the sample numbers, MOFA is an unsupervised factor-analysis framework. It aims to represent the integrated datasets using a smaller number of latent factors. Each factor captures a source of variation that may be specific to one omics view or shared across multiple views. In this practical, the factors are inferred jointly from the RNA-seq and methylation data rather than being constructed to predict the Low, Medium, and High boar-taint classes.

The parameter `model_opts$num_factors <- 3` requests the estimation of three latent factors. These factors provide three independent dimensions through which the principal patterns of variability in the RNA-seq and methylation datasets can be summarized. The number of factors is specified before training and should be interpreted in relation to the amount of variance explained by each factor during downstream analysis. The MOFA documentation states that there is no universally optimal number of factors because the appropriate choice depends on the objective, dimensions, and complexity of the analysis.

The `prepare_mofa()` function combines the selected data, model, and training options with the initial MOFA object. This preparation step checks and configures the object before model fitting. The sample metadata and groups are not supplied again because they were already incorporated when `create_mofa()` was called. Finally, `run_mofa()` trains the prepared model. The output file path is defined as `model.hdf5` within the current working directory. The HDF5 file stores the trained model and its learned parameters. Setting `use_basilisk = TRUE` instructs MOFA2 to use the Python environment managed through `basilisk`. For a clearer understanding of the integration, we invite you to look at the general framework in: `https://biofam.github.io/MOFA2/`.

```{r mofa-training}
# Create MOFA object with groups specified explicitly - it now access the list 
mofa_object <- create_mofa(multi_omics_data, groups = sample_metadata$group)
plot_data_overview(mofa_object)

# Prepare options
data_opts <- get_default_data_options(mofa_object)
model_opts <- get_default_model_options(mofa_object)
train_opts <- get_default_training_options(mofa_object)
train_opts$seed <- 42  # for reproducibility
train_opts$maxiter <- 1000
model_opts$num_factors <- 3

# Prepare the model (NO samples_metadata or groups arguments here)
mofa_object <- prepare_mofa(
  object = mofa_object,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

# Run MOFA
outfile = file.path(getwd(),"model.hdf5")
mofa_trained <- run_mofa(mofa_object,outfile, use_basilisk = TRUE)

```

------------------------------------------------------------------------

# 8. Inspect variance explained by the trained model

The first step in the MOFA analysis is to quantify the amount of variance explained (R2) by each factor in each data modality. The variance explained estimates are stored in the hdf5 file and loaded in `mofa_trained@cache$variance_explained$r2_per_factor[[1]]`. As observed, RNA-seq liver data has an important cumulative variance explained of 52 % and methylation data of 32.43 % among the three observed factors. We will inspect in the future cells the biomarkers driven by the Views (Modalities) across groups differentiation.

```{r variance-explained}
# -----------------------------------
# Downstream Analysis & Visualization
# -----------------------------------

### Check the Number of features according to groups
plot_data_overview(mofa_trained)
# Total variance explained per Omics 
print(mofa_trained@cache$variance_explained$r2_total[[1]])
# Variance explained by each factor per view 
head(mofa_trained@cache$variance_explained$r2_per_factor[[1]])
###Plot variance 
plot_variance_explained(mofa_trained, x="view", y="factor") # Modality
plot_variance_explained(mofa_trained, x="group", y="factor", plot_total = T)[[2]] # Group


```

------------------------------------------------------------------------

# 9. Plot the MOFA factors by group

<!-- Add your comments here. -->

```{r factor-plot}
plot_factor(mofa_trained, 
                 factors = c(1,2),
                 color_by = "group",
                 dot_size = 3,        # change dot size
                 dodge = T,           # dodge points with different colors
                 legend = T,         
                 add_violin = T,      
                 violin_alpha = 0.25)



```

------------------------------------------------------------------------

# 10. Plot Factor 1 against Factor 2

<!-- Add your comments here. -->

```{r factor-scatter}
# factors scatter plot
# Convert long to wide format
factors_df <- get_factors(mofa_trained, factors = "all", as.data.frame = TRUE)

factors_wide <- factors_df %>%
  pivot_wider(names_from = factor, values_from = value)

# Scatter plot
ggplot(factors_wide, aes(x = Factor1, y = Factor2, color = group)) +
  geom_point(size = 3) +
  geom_text(aes(label = sample), vjust = -0.5, hjust = 0.5, size = 3) +  # sample labels above points
  theme_minimal() +
  labs(title = "MOFA Factor 1 vs Factor 2 by Group", color = "Group")



```

------------------------------------------------------------------------

# 11. Inspect RNA-seq feature weights

<!-- Add your comments here. -->

```{r rna-feature-weights}
################

# Identify Important Genes/CpGs for Each Factor
plot_weights(mofa_trained,
             view = "RNAseq",
             factor = 1,
             nfeatures = 10,     # Number of features to highlight
             scale = T,          
             abs = F             # Take the absolute value?
)

# Plot top absolute weights (most influential features regardless of direction)
plot_top_weights(mofa_trained,
                 view = "RNAseq",
                 factor = 1,
                 nfeatures = 10
)


```

------------------------------------------------------------------------

# 12. Inspect methylation feature weights

<!-- Add your comments here. -->

```{r methylation-feature-weights}
#### Methylation 
# Plot feature weights for Factor 1 in the Methylation view
plot_weights(mofa_trained,
             view = "Methylation",
             factor = 1,
             nfeatures = 10,     # Number of features to highlight
             scale = T,          # Scale weights from -1 to 1
             abs = F             # Take the absolute value?
)

# Plot top absolute weights (most influential features regardless of direction)
plot_top_weights(mofa_trained,
                 view = "Methylation",
                 factor = 1,
                 nfeatures = 10
)


```

------------------------------------------------------------------------

# 13. Relate the top RNA-seq feature to Factor 1 and Factor 2

<!-- Add your comments here. -->

```{r factor-feature-correlation}
#######  Gene correlates with factor 1 component ###############################

# Extract weights
weights <- get_weights(mofa_trained, views = "RNAseq", factors = 1, as.data.frame = TRUE)
#These weights represent how strongly each feature contributes to Factor 1

top_genes <- get_weights(mofa_trained, 
                         view = "RNAseq", 
                         factor = 1, 
                         as.data.frame = TRUE)

top_feature <- top_genes[order(abs(top_genes$value), decreasing = TRUE), ]
top_feature <- weights[order(-abs(weights$value)), "feature"][1]
head(top_feature)
# Extract factor scores
factors <- get_factors(mofa_trained, factors = 1, as.data.frame = TRUE)

# Extract feature expression
data <- get_data(mofa_trained, view = "RNAseq", as.data.frame = TRUE)
feature_data <- subset(data, as.character(feature) == as.character(top_feature))

# Merge data
df <- merge(factors, feature_data, by = "sample")
df$group <- samples_metadata(mofa_trained)$group

# Plot 
ggplot(df, aes(x = value.x, y = value.y, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  stat_cor(method = "pearson") +
  labs(
    x = "Factor 1 score",
    y = paste("Expression of", top_feature)
  ) +
  theme_minimal()



```

```{r factor-feature-correlation}
#######  Gene correlates with factor 1 component ###############################

# Extract weights
weights <- get_weights(mofa_trained, views = "RNAseq", factors = 2, as.data.frame = TRUE)
#These weights represent how strongly each feature contributes to Factor 1

top_genes <- get_weights(mofa_trained, 
                         view = "RNAseq", 
                         factor = 2, 
                         as.data.frame = TRUE)

top_feature <- top_genes[order(abs(top_genes$value), decreasing = TRUE), ]
top_feature <- weights[order(-abs(weights$value)), "feature"][1]
head(top_feature)
# Extract factor scores
factors <- get_factors(mofa_trained, factors = 2, as.data.frame = TRUE)

# Extract feature expression
data <- get_data(mofa_trained, view = "RNAseq", as.data.frame = TRUE)
feature_data <- subset(data, as.character(feature) == as.character(top_feature))

# Merge data
df <- merge(factors, feature_data, by = "sample")
df$group <- samples_metadata(mofa_trained)$group

# Plot 
ggplot(df, aes(x = value.x, y = value.y, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  stat_cor(method = "pearson") +
  labs(
    x = "Factor 1 score",
    y = paste("Expression of", top_feature)
  ) +
  theme_minimal()



```

------------------------------------------------------------------------

# 14. Visualize the top features driving Factor 1

<!-- Add your comments here. -->

```{r top-factor-features}
# Visualization of Top 30 Features Driving Latent Factor 1 in MOFA (RNA-Seq and Methylation)
# Get latent factors (samples × factors)
factors <- get_factors(mofa_trained, factors = "all")

# Get feature weights for all views (features × factors)
weights <- get_weights(mofa_trained, views = c("RNAseq", "Methylation"), factors = 1)

# Top 30 genes with highest absolute weights in RNAseq
top_genes <- sort(abs(weights$RNAseq[,1]), decreasing = TRUE)[1:30]
top_genes_names <- names(top_genes)

# Top 30 CpGs with highest absolute weights in Methylation
top_cpgs <- sort(abs(weights$Methylation[,1]), decreasing = TRUE)[1:30]
top_cpgs_names <- names(top_cpgs)

# Get input data (long format)
data <- get_data(mofa_trained, as.data.frame = TRUE)

# Subset data for top features
top_gene_data <- subset(data, feature %in% top_genes_names & view == "RNAseq")
top_cpg_data <- subset(data, feature %in% top_cpgs_names & view == "Methylation")


# Plot Genes
ggplot(top_gene_data, aes(x = group, y = value, fill = group)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.2, size = 1) +
  facet_wrap(~ feature, scales = "free_y") +
  theme_minimal() +
  labs(title = "Top 30 Genes by MOFA Factor 1",
       y = "Expression", x = "Boar Taint Group")

##### For CPGs 
ggplot(top_cpg_data, aes(x = group, y = value, fill = group)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.2, size = 1) +
  facet_wrap(~ feature, scales = "free_y") +
  theme_minimal() +
  labs(title = "Top 30 CPGs by MOFA Factor 1",
       y = "Expression", x = "Boar Taint Group")

```
