---
title: "Multi-Omics Integration Analysis - MOFA"
author: "Adam El Idrissi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Practical overview

This R Markdown document preserves the workflow and execution order of the supplied MOFA.R script, applied in the same dataset we analysed previously. This document contains detailed descriptions of the code cells and the resulting outputs in relation to the dataset used in this practical. It provides an overview of the multi-omics integration process performed using `Multi-Omics Factor Analysis (MOFA)`. The official MOFA2 tutorials, available at `https://biofam.github.io/MOFA2/tutorials.html`, which provides additional examples of model training and downstream analysis.

MOFA utilizes Matrix Factorization framework for the unsupervised integration of multiple omics datasets. It learns an interpretable, low-dimensional representation of the integrated data through a set of latent factors. These factors capture the major sources of variation across the analysed omics modalities and can subsequently be examined in relation to sample groups, biological characteristics, and influential molecular features.

------------------------------------------------------------------------

# 1. Package installation and loading

<!-- Add your comments here. -->

```{r package-loading}


########## MOFA ################################################################

## Load all the required libraries 
# BiocManager::install("MOFA2")


library(MOFA2)
library(data.table)
library(ggplot2)
library(dplyr)
library(psych)
library(GGally)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(basilisk)
library(edgeR)

```

------------------------------------------------------------------------

# 2. Load the RNA-seq, methylation, and metadata files

<!-- Add your comments here. -->

```{r data-import}
##### load all the required files and arrange the files for MOFA Analysis
RNA_Seq_Data=read.csv("RNA_Seq_Liver.csv", row.names = 1)
head(RNA_Seq_Data)[1:5]


Methylation=read.csv("Epigenetic_matrix_filtered_100000.csv", row.names = 1)
rownames(Methylation)=Methylation$region
Methylation$X=NULL
Methylation$region =NULL

head(Methylation)

metadat_columns=read.csv("Metadata.csv")
head(metadat_columns)
metadat_columns$X=NULL

```

------------------------------------------------------------------------

# 3. Metadata filtering and sample alignment

<!-- Add your comments here. -->

```{r metadata-filtering}
###################### Metadata Filtering ######################################
# Add prefix based on boar taint group
metadat_columns$SampleName <- ifelse(
  grepl("high", metadat_columns$Group, ignore.case = TRUE), 
  paste0("H", metadat_columns$SampleID),
  ifelse(
    grepl("medium", metadat_columns$Group, ignore.case = TRUE), 
    paste0("M", metadat_columns$SampleID),
    ifelse(
      grepl("low", metadat_columns$Group, ignore.case = TRUE), 
      paste0("L", metadat_columns$SampleID),
      paste0("X", metadat_columns$SampleID)  # fallback if none match
    )
  )
)
metadat_columns$Group <- sub(" .*", "", metadat_columns$Group)



# Create the old and new column name mapping
metadat_columns$OldSampleName <- paste0("X", metadat_columns$SampleID)  # Original names
name_map <- setNames(metadat_columns$SampleName, metadat_columns$OldSampleName)  # Named vector: old -> new
# Rename methylation matrix columns
common_meth_cols <- intersect(colnames(Methylation), names(name_map))
colnames(Methylation)[match(common_meth_cols, colnames(Methylation))] <- name_map[common_meth_cols]
# Rename RNA-seq matrix columns
common_rna_cols <- intersect(colnames(RNA_Seq_Data), names(name_map))
colnames(RNA_Seq_Data)[match(common_rna_cols, colnames(RNA_Seq_Data))] <- name_map[common_rna_cols]
# Update metadata rownames to match new names
rownames(metadat_columns) <- metadat_columns$SampleName
# Filter metadata to keep only samples present in your omics data
omics_samples <- colnames(Methylation)  
sample_metadata <- metadat_columns[metadat_columns$SampleName %in% omics_samples, ]


# Double-check that sample orders match
all(colnames(RNA_Seq_Data) == colnames(Methylation))
# Reorder columns of methylation matrix to match RNA-seq
meth_matrix <- Methylation[, colnames(RNA_Seq_Data)]
####### Check that they're now aligned
all(colnames(meth_matrix) == colnames(RNA_Seq_Data))  # should return TRUE

## Change the colnames of metadata
colnames(sample_metadata)[colnames(sample_metadata) == "SampleName"] <- "sample"
colnames(sample_metadata)[colnames(sample_metadata) == "Group"] <- "group"

```

------------------------------------------------------------------------

# 4. Normalize and filter the RNA-seq data

<!-- Add your comments here. -->

```{r rna-normalization}
########### Normalize and filter for downstream analysis #######################

# Create DGEList object
dge <- DGEList(counts = RNA_Seq_Data)
keep <- filterByExpr(dge)  
dge <- dge[keep, , keep.lib.sizes=FALSE]
# Normalize with TMM
dge <- calcNormFactors(dge)
# Log2-CPM transformation
log_RNA <- cpm(dge, log = TRUE, prior.count = 1)
# Filter top 3000 most variable genes
gene_var <- apply(log_RNA, 1, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:3000]
log_RNA_filtered <- log_RNA[top_genes, ]


```

------------------------------------------------------------------------

# 5. Transform and filter the methylation data

<!-- Add your comments here. -->

```{r methylation-filtering}
### Methylation data
# Ensure beta values are in [0, 1]
Methylation[Methylation <= 0] <- 0
Methylation[Methylation >= 1] <- 1 
summary(Methylation)
# Convert data frame to numeric matrix
Methylation <- as.matrix(Methylation)
summary(Methylation)
storage.mode(Methylation) <- "numeric"  # make sure all values are numeric
# Convert to M-values
Mval <- log2(Methylation / (1 - Methylation))
# Replace Inf and -Inf with NA safely
Mval[is.infinite(Mval)] <- NA
class(Mval)
# keep features with <20% NA
clean_meth <- Mval[apply(Mval, 1, function(x) sum(is.na(x)) < length(x) * 0.2), ]  
clean_meth <- clean_meth[, colSums(is.na(clean_meth)) < nrow(clean_meth) * 0.2]

# Calculate variance per CpG (row), ignoring NAs
var_per_cpg <- apply(clean_meth, 1, var, na.rm = TRUE)
# Sort CpGs by variance decreasing and get top 5000 CpGs
top_CPGs <- names(sort(var_per_cpg, decreasing = TRUE))[1:5000]
# Subset methylation data to top variable CpGs
Meth_view <- clean_meth[top_CPGs, ]
# Check dimensions
dim(Meth_view)
# Confirm shared sample names
common_samples <- intersect(colnames(log_RNA_filtered), colnames(Meth_view))



```

------------------------------------------------------------------------

# 6. Prepare the multi-omics data views

<!-- Add your comments here. -->

```{r multi-omics-data}
# Subset and reorder both datasets to same sample order
RNA_view <- log_RNA_filtered[, common_samples]
Meth_view <- Meth_view[, common_samples]



# Build multi-omics data list (9 sample names in columns, rows have RNASeq in one object, 
# similarly in a separate object another list with the CpG data)- They are NOT merged in a single object or data frame.
multi_omics_data <- list(
  RNAseq = as.matrix(RNA_view),
  Methylation = as.matrix(Meth_view)
)


```

------------------------------------------------------------------------

# 7. Create, prepare, and train the MOFA model

<!-- Add your comments here. -->

```{r mofa-training}
# Create MOFA object with groups specified explicitly - it now access the list 
mofa_object <- create_mofa(multi_omics_data, groups = sample_metadata$group)
plot_data_overview(mofa_object)

# Prepare options
data_opts <- get_default_data_options(mofa_object)
model_opts <- get_default_model_options(mofa_object)
train_opts <- get_default_training_options(mofa_object)
train_opts$seed <- 42  # for reproducibility
train_opts$maxiter <- 1000
model_opts$num_factors <- 2

# Prepare the model (NO samples_metadata or groups arguments here)
mofa_object <- prepare_mofa(
  object = mofa_object,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

# Run MOFA
outfile = file.path(getwd(),"model.hdf5")
mofa_trained <- run_mofa(mofa_object,outfile, use_basilisk = TRUE)

```

------------------------------------------------------------------------

# 8. Inspect variance explained by the trained model

<!-- Add your comments here. -->

```{r variance-explained}
# -----------------------------------
# Downstream Analysis & Visualization
# -----------------------------------

### Check the Number of features according to groups
plot_data_overview(mofa_trained)
# Total variance explained per Omics 
print(mofa_trained@cache$variance_explained$r2_total[[1]])
# Variance explained by each factor per view 
head(mofa_trained@cache$variance_explained$r2_per_factor[[1]])
###Plot variance 
plot_variance_explained(mofa_trained, x="view", y="factor")
plot_variance_explained(mofa_trained, x="group", y="factor", plot_total = T)[[2]]


```

------------------------------------------------------------------------

# 9. Plot the MOFA factors by group

<!-- Add your comments here. -->

```{r factor-plot}
plot_factor(mofa_trained, 
                 factors = c(1,2),
                 color_by = "group",
                 dot_size = 3,        # change dot size
                 dodge = T,           # dodge points with different colors
                 legend = T,         
                 add_violin = T,      
                 violin_alpha = 0.25)




```

------------------------------------------------------------------------

# 10. Plot Factor 1 against Factor 2

<!-- Add your comments here. -->

```{r factor-scatter}
# factors scatter plot
# Convert long to wide format
factors_df <- get_factors(mofa_trained, factors = "all", as.data.frame = TRUE)

factors_wide <- factors_df %>%
  pivot_wider(names_from = factor, values_from = value)

# Scatter plot
ggplot(factors_wide, aes(x = Factor1, y = Factor2, color = group)) +
  geom_point(size = 3) +
  geom_text(aes(label = sample), vjust = -0.5, hjust = 0.5, size = 3) +  # sample labels above points
  theme_minimal() +
  labs(title = "MOFA Factor 1 vs Factor 2 by Group", color = "Group")



```

------------------------------------------------------------------------

# 11. Inspect RNA-seq feature weights

<!-- Add your comments here. -->

```{r rna-feature-weights}
################

# Identify Important Genes/CpGs for Each Factor
plot_weights(mofa_trained,
             view = "RNAseq",
             factor = 1,
             nfeatures = 10,     # Number of features to highlight
             scale = T,          
             abs = F             # Take the absolute value?
)

# Plot top absolute weights (most influential features regardless of direction)
plot_top_weights(mofa_trained,
                 view = "RNAseq",
                 factor = 1,
                 nfeatures = 10
)


```

------------------------------------------------------------------------

# 12. Inspect methylation feature weights

<!-- Add your comments here. -->

```{r methylation-feature-weights}
#### Methylation 
# Plot feature weights for Factor 1 in the Methylation view
plot_weights(mofa_trained,
             view = "Methylation",
             factor = 1,
             nfeatures = 10,     # Number of features to highlight
             scale = T,          # Scale weights from -1 to 1
             abs = F             # Take the absolute value?
)

# Plot top absolute weights (most influential features regardless of direction)
plot_top_weights(mofa_trained,
                 view = "Methylation",
                 factor = 1,
                 nfeatures = 10
)


```

------------------------------------------------------------------------

# 13. Relate the top RNA-seq feature to Factor 1

<!-- Add your comments here. -->

```{r factor-feature-correlation}
#######  Gene correlates with factor 1 component ###############################

# Extract weights
weights <- get_weights(mofa_trained, views = "RNAseq", factors = 1, as.data.frame = TRUE)
#These weights represent how strongly each feature contributes to Factor 1

top_genes <- get_weights(mofa_trained, 
                         view = "RNAseq", 
                         factor = 1, 
                         as.data.frame = TRUE)

top_feature <- top_genes[order(abs(top_genes$value), decreasing = TRUE), ]
top_feature <- weights[order(-abs(weights$value)), "feature"][1]
head(top_feature)
# Extract factor scores
factors <- get_factors(mofa_trained, factors = 1, as.data.frame = TRUE)

# Extract feature expression
data <- get_data(mofa_trained, view = "RNAseq", as.data.frame = TRUE)
feature_data <- subset(data, as.character(feature) == as.character(top_feature))

# Merge data
df <- merge(factors, feature_data, by = "sample")
df$group <- samples_metadata(mofa_trained)$group

# Plot 
ggplot(df, aes(x = value.x, y = value.y, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  stat_cor(method = "pearson") +
  labs(
    x = "Factor 1 score",
    y = paste("Expression of", top_feature)
  ) +
  theme_minimal()



```

------------------------------------------------------------------------

# 14. Visualize the top features driving Factor 1

<!-- Add your comments here. -->

```{r top-factor-features}
# Visualization of Top 30 Features Driving Latent Factor 1 in MOFA (RNA-Seq and Methylation)
# Get latent factors (samples × factors)
factors <- get_factors(mofa_trained, factors = "all")

# Get feature weights for all views (features × factors)
weights <- get_weights(mofa_trained, views = c("RNAseq", "Methylation"), factors = 1)

# Top 30 genes with highest absolute weights in RNAseq
top_genes <- sort(abs(weights$RNAseq[,1]), decreasing = TRUE)[1:30]
top_genes_names <- names(top_genes)

# Top 30 CpGs with highest absolute weights in Methylation
top_cpgs <- sort(abs(weights$Methylation[,1]), decreasing = TRUE)[1:30]
top_cpgs_names <- names(top_cpgs)

# Get input data (long format)
data <- get_data(mofa_trained, as.data.frame = TRUE)

# Subset data for top features
top_gene_data <- subset(data, feature %in% top_genes_names & view == "RNAseq")
top_cpg_data <- subset(data, feature %in% top_cpgs_names & view == "Methylation")


# Plot Genes
ggplot(top_gene_data, aes(x = group, y = value, fill = group)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.2, size = 1) +
  facet_wrap(~ feature, scales = "free_y") +
  theme_minimal() +
  labs(title = "Top 30 Genes by MOFA Factor 1",
       y = "Expression", x = "Boar Taint Group")

##### For CPGs 
ggplot(top_cpg_data, aes(x = group, y = value, fill = group)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.2, size = 1) +
  facet_wrap(~ feature, scales = "free_y") +
  theme_minimal() +
  labs(title = "Top 30 CPGs by MOFA Factor 1",
       y = "Expression", x = "Boar Taint Group")

```
